Expert-In-The-Loop Causal Discovery: Iterative Model Refinement Using Expert Knowledge

### 2025-02-10 paper/main.tex
Updates

\begin{figure}[t!]
	\begin{subfigure}{0.25\textwidth}
		\centering
		\includegraphics{../code/fig4/unexplained_effect.pdf}
		\caption{}
	\end{subfigure}%
	\begin{subfigure}{0.25\textwidth}
		\centering
		\includegraphics{../code/fig4/ll.pdf}
		\caption{}
	\end{subfigure}
	\caption{Plots showing improvements in fit of the model over iterations
	on the Adult income dataset. For the expert simulation an LLM is used
	for the edge orientations. (a) Shows using total residual association that is the
	sum of effect size between variables that don't have an edge between them. (b)
	Shows the log-likelihood of the overall model.
	}
	\label{fig:unexplained_ll}
\end{figure}

@@

\begin{equation}
	\begin{split}
		\beta &= \mathcal{N}(\bm{0}, \bm{I}) \\
		Y_i &= \bm{\beta} \mathrm{Pa}_G(Y_i) + \mathcal{N}(0, 1)
	\end{split}
\end{equation}

@@

Our procedure closely resembles certain score-based automated causal discovery
methods, such as Greedy Equivalence Search (GES), where the algorithm
iteratively adds or removes edges that improve the scoring metric the most.
Similarly, a greedy version of our approach would involve adding an edge
between the pair of variables with the highest residual association. While
both methods aim to find modifications that most improve the model at each
step, a key distinction lies in the nature of the evaluation criteria: unlike
standard scoring metrics, our measure of association is not decomposable. A key
property of scoring metrics is that they should be decomposible, i.e., they can
be expressed as the sum of scores of nodes given its parents. This means adding
or removing an edge affects only a localized part of the model. In contrast,
our measure of association is global, meaning that modifying an edge in one
part of the model can influence association values elsewhere.

Given a DAG $ G = (V, E) $, the total residual association, $\tau$ in define as:
\begin{equation}
	\tau = \sum_{\substack{X, Y \in V \\ X \rightarrow Y, Y \rightarrow X \not \in E}}   \phi(X, Y, \mathrm{pa}_G(X) \cup \mathrm{pa}_G(Y))
\end{equation}

Another major difference is in the interpretability of the evaluation metric.
Most scoring metrics are based on log-likelihood with a penalty for model
complexity. These score metrics allow for relative comparisons between models
but do not provide an absolute measure of model fit. That is, they indicate
which model is better for a given dataset but do not reveal how well the model
explains the data in an absolute sense. In contrast, summing our measure of
association across all variables provides a direct fitness measure—this sum
approaches zero when the model perfectly accounts for all observed correlations
in the data. This property enables an absolute assessment of model quality,
rather than just a relative comparison between models. Figure~\ref{fig:unexplained_ll}

>>>>

\begin{figure}[t!]
	\begin{subfigure}{0.25\textwidth}
		\centering
		\includegraphics{../code/fig4/unexplained_effect.pdf}
		\caption{}
	\end{subfigure}%
	\begin{subfigure}{0.25\textwidth}
		\centering
		\includegraphics{../code/fig4/ll.pdf}
		\caption{}
	\end{subfigure}
	\caption{Model fit improvement over iterations on the Adult Income dataset. In the expert simulation, an 
		LLM is used to determine the edge orientations. (a) Shows the total residual association, $\tau$. 
		(b) Shows the log-likelihood of the overall model.
	}
	\label{fig:unexplained_ll}
\end{figure}

@@

\begin{equation}
	\begin{split}
		\beta &= \mathrm{Uniform}([-0.6, 0.6]) \\
		Y_i &= \bm{\beta} \mathrm{Pa}_G(Y_i) + \mathcal{N}(0, 1)
	\end{split}
\end{equation}

@@

Our procedure shares similarities with score-based automated causal discovery
methods, such as Greedy Equivalence Search (GES), where the algorithm
iteratively adds or removes edges that maximize improvements in a specified
scoring metric. Given a scoring metric, --total residual association, $ \tau $ for 
a DAG $ G = (V, E) $, defined as:

\begin{equation}
	\tau = \sum_{\substack{X, Y \in V \\ X \rightarrow Y, Y \rightarrow X \not \in E}}   \phi(X, Y, \mathrm{pa}_G(X) \cup \mathrm{pa}_G(Y))
\end{equation}

our \textsc{RankedExpand} approach behaves similarly to GES. In each iteration,
we add the edge between the variable pair that results in the lowest $ \tau $
and remove edges that have no impact on $ \tau $.

While both methods aim to make iterative modifications to the DAG, a key
distinction lies in the nature of the evaluation criteria: unlike standard
scoring metrics, $ \tau $ is not decomposable. Decomposibility is a key
property of scoring metrics, allowing them to be expressed as the sum of
individual variables in the model. This property ensures that adding or
removing an edge only affects a localized part of the model. In contrast, $
\tau $ can not be decomposed, as modifying an edge in one part of the model can
affect the residual associations in other parts of the model.

Another key difference lies in the interpretability of the evaluation metric.
Scoring metrics are based on usually based on the log-likelihood with a penalty
for model complexity. They can be used to make relative comparisons between
models but their values do not have any interpretation in an absolute sense.
That is, they indicate which model is better for a given dataset but do not
quantify how well the model explains the data in an absolute sense. In
contrast, $ \tau $ provides an absolute measure of model fit -- its value
approaches $ 0 $ as the model perfectly explains the observed data.
Figure~\ref{fig:unexplained_ll} illustrates this distinction by comparing GES
to \textsc{ExpertInLoop} on the Adult income dataset across iterations of a
single run of the algorithms.
